\large How resource abundance and stochasticity affect animals' spatial needs

### 2022-11-17 writing/manuscript.Rmd
resolve conflict

@@

\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@baldwin_nutritional_1984; @reich_body_2001; @ref]. Failing to acquire sufficient resources is likely to result in lower fitness, changes in physiology or behavior, physiological damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism [@ref] and movement costs [@ref]. The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [see figure \@ref(fig:mu-hyp) and @nilsen_can_2005; @ex]. Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan \@ref(fig:mu-hyp). As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].

@@

This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize. Such a model may be appropriate in regions with little to no changes in variability and where unpredictable events occur at a similar frequency over time, or when data availability is too low to produce appreciable measures of variance. Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts). When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is rarely the case [@refs; but see @refs]. Most often, $\e(R)$ will vary over time, and animals' behaviors will change in response. Additionally, when data availability is sufficiently high to produce precise estimates of $\var(R)$, the effect of $\var(R)$ on animals' $\vpos$ should be estimated before assuming the effect is negligible. The need to account for $\var(R)$ is compounded by recent changes in climate, which exposes species to novel situations and increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018]. Additionally, anthropogenic structures and urban areas reduce the habitat available to many terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018]. Although some species rely on anthropogenic areas and structures for predicable and dependable sources of resources [steppe rat snake: @yu_artificial_2022; yellow mongoose: @cronk_home_2021; common octopus: @arechavala-lopez_common_2019; raccoon dog: @mitsuhashi_home_2018; coyote: @peron_periodic_2017], anthropogenic environments can be highly unpredictable [@ref] and pose high mortality risks [@noonan_roads_2021; @refs].

@@

<!--
in stochastic environments:
- higher spatial needs and possibly store resources for periods of scarcity, which are expected to get longer with climate change
- heterogeneous environments, such as habitats with a high variety of patches, may force animals to move between patches if they cannot depend on a single one
-->

@@

@@

```{r moving-window-examples, fig.cap="Seven-day home range size of a tapir ($Tapirus~terrestris$) in response to changes in mean and variance in resource abundance. (a.) GPS tracking data of the tapir. (b.) Mean resource abundance esimated as the mean NDVI at the locations visited by the tapir. (c.) Varince in resource abundance esimated as the average variance in NDVI at the locations visited by the tapir. (d.) Estimated home range size during each seven-day period, based on 50\\% (bold) and 95\\% (thin) utilization quantiles. (e.) Effect of resource abundance on home range size. (f.) Effect of resource unpredictability on home range size. The effects in panels e and f were estimated using independent generalized linear models with Gamma conditional distributions. The tapir movement data corresponds to the individual named \"Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}
knitr::include_graphics('../figures/tapir-example.png')
```

>>>>

***Not sure if we need this or if we can tie it in somewhere:***

@@

\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@baldwin_nutritional_1984; @reich_body_2001; @ref]. Failing to acquire sufficient resources is likely to result in lower fitness, changes in physiology or behavior, physiological damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism [@ref] and movement costs [@ref]. The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not linear [see figure \@ref(fig:mu-hyp) and @nilsen_can_2005; @ex]. Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan \@ref(fig:mu-hyp). As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].

@@

This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize. Such a model may be appropriate in regions with little to no changes in variability and where unpredictable events occur at a similar frequency over time, or when data availability is too low to produce appreciable measures of variance. Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts). When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is rarely the case. Most often, $\e(R)$ will vary over time, and animals' behaviors will change in response. Additionally, when data availability is sufficiently high to produce precise estimates of $\var(R)$, the effect of $\var(R)$ on animals' $\vpos$ should be estimated before assuming the effect is negligible. The need to account for $\var(R)$ is compounded by recent changes in climate, which exposes species to novel situations and increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018]. Additionally, anthropogenic structures and urban areas reduce the habitat available to many terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018]. Although some species rely on anthropogenic areas and structures for predicable and dependable sources of resources [steppe rat snake: @yu_artificial_2022; yellow mongoose: @cronk_home_2021; common octopus: @arechavala-lopez_common_2019; raccoon dog: @mitsuhashi_home_2018; coyote: @peron_periodic_2017], anthropogenic environments can be highly unpredictable [@ref] and pose high mortality risks [@noonan_roads_2021; @refs].

@@

Figure \@ref(fig:5-5-sims) illustrates the simulated average spatial requirements of an animal as a function of $\e(R)$ and $\var(R)$. The top row (constant $\var(R)$) shows how $\vpos$ changes as a function of $\e(R)$ while $\var(R)$ remains constant. (Some noise in $H$ is still present since $\var(R) \ne 0$ and thus $R$ is not constant). Note the nonlinear decrease in $H$ as $\e(R)$ increases at a constant slope (linear $\e(R)$), with larger changes when $\e(H)$ is low. This is because when $\e(R)$ is low, small additive changes in $\e(R)$ (e.g. from 1 to 2) can correspond to large multiplicative changes (i.e., 2 is twice as big as 1). Additionally, the 95% home range is more sensitive to changes in $\e(R)$ than the core home range $H_{50\%}$, since (not sure about this) larger quantiles (e.g., 95% > 50%) are more sensitive to changes in a distribution's variance (in this case $\propto \vpos$).

@@

The leftmost column of figure \@ref(fig:5-5-sims) (constant $\e(R)$) illustrates how an animal's home range may vary as $\var(R)$ changes over time while $\e(R)$ remains constant. Overall, $\vpos$ increases with $\var(R)$, although the 95% home range ($H_{95\%}$) is more sensitive to changes in $\var(R)$ than the 50% core home range ($H_{50\%}$). ***here*** <!--Additionally, the effect of $\var(R)$ on $H$ depends on $\e(R)$, since the effect of $\var(R)$ on $H$ is stronger when $\e(R)$ is low (see the bottom row with erratic $\var(R)$).-->

@@

```{r mw-examples, fig.cap="Seven-day home range size of a tapir ($Tapirus~terrestris$) in response to changes in mean and variance in resource abundance. (a.) GPS tracking data of the tapir. (b.) Mean resource abundance esimated as the mean NDVI at the locations visited by the tapir. (c.) Varince in resource abundance esimated as the average variance in NDVI at the locations visited by the tapir. (d.) Estimated home range size during each seven-day period, based on 50\\% (bold) and 95\\% (thin) utilization quantiles. (e.) Effect of resource abundance on home range size. (f.) Effect of resource unpredictability on home range size. The effects in panels e and f were estimated using two generalized linear models with Gamma families of distributions. The tapir movement data corresponds to the individual named \"Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}
knitr::include_graphics('../figures/tapir-example.png')
```
